LabMate: a prospectus for types for MATLAB

### 2024-04-10 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] typo
[ writing ] section 2 wip

\title{LabMate: supporting types for MATLAB}
\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
\affiliation{%
\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , UK}  \\
\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}}

@@

\begin{abstract}
  Many computations in science and engineering are implemented in the
  programming language MATLAB. However the high-level meaning of such
  MATLAB programs stay informal, which can lead to implementation
  errors and bugs, for example relating to incompatible units of
  measure for quantities, or incompatible sizes of matrices at
  runtime. We are in the process of developing LabMate, which is a
  tool for reifying current informal programmer practices into a
  language of formal comments. These comments are ignored by MATLAB,
  but acted on and checked by LabMate. We outline the design
  principles behind LabMate, our current progress, and our future
  plans.
\end{abstract}
%\begin{keywords}
%  MATLAB, software correctness, dimensional consistency, type theory
%\end{keywords}

@@

\[
\begin{pmatrix*}
  d_{\textrm{nc}1} & d_{\textrm{c}1} & n_1 & c_1 & t_1  \\
  d_{\textrm{nc}2} & d_{\textrm{c}2} & n_2 & c_2 & t_2  \\
   & \vdots &  & & \\
  d_{\textrm{nc}N} & d_{\textrm{c}N} & n_N & c_N & t_N  \\
\end{pmatrix*}
\times
\begin{pmatrix}
  a \\
  a_{cal} \\
  c \\
  c_{cal} \\
  m
\end{pmatrix}
=
\begin{pmatrix}
  v_1 \\
  v_2 \\
  \vdots \\
  v_N
\end{pmatrix}
\]
%
The matrix on the left is made up of the following values:
%
\begin{itemize}
    \item The values in the first four columns are purely numeric and have no associated dimension.

@@

    \item The forth column is the opposite of the third.

@@

The first row vector contains values to be calculated using the line fit. The second column vector contains voltage readings. The measured value of the resistor is calculated using the contents of the first row vector. Details of that calculation are beyond the scope of this paper.

This calculation can be converted into a MATLAB problem with LabMate annotations as follows, using MATLAB's builtin left division operator \texttt{A \textbackslash{} b} to compute a least squares fit. We then ask for the type of the resulting solution. Note that for reasons of space only twelve values are listed. A real measurement would contain considerably more values:
\begin{verbatim}
%> A :: [ 12 x 5 ] double
A = [ 1  0  1  0  t1
      1  0  1  0  t2
      1  0  1  0  t3
     -1  0  1  0  t4
     -1  0  1  0  t5
     -1  0  1  0  t6
      0  1  0  1  t7
      0  1  0  1  t8
      0  1  0  1  t9
      0 -1  0  1  t10
      0 -1  0  1  t11
      0 -1  0  1  t12 ]

@@

Did you notice the error in the above code? The vector \texttt{b} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate detects this error and produces the following as output.
\begin{verbatim}
%> A :: [ 12 x 5 ] double
A = [ 1  0  1  0  t1
      1  0  1  0  t2
      1  0  1  0  t3
     -1  0  1  0  t4
     -1  0  1  0  t5
     -1  0  1  0  t6
      0  1  0  1  t7
      0  1  0  1  t8
      0  1  0  1  t9
      0 -1  0  1  t10
      0 -1  0  1  t11
      0 -1  0  1  t12 ]

@@

x = A \ B

>>>>

\title{LabMate: a prospectus for types for MATLAB}
\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Andr\'e Videla  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
\affiliation{%
\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , UK}  \\
\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}}

@@

\begin{abstract}
  Many computations in science and engineering are implemented in the
  programming language MATLAB. However the high-level meaning of such
  MATLAB programs stay informal, which can lead to implementation
  errors and bugs, for example relating to incompatible units of
  measure for quantities, or incompatible sizes of matrices at
  runtime. We are in the process of developing LabMate, which is a
  tool for reifying current informal programmer practices into a
  language of formal comments. These comments are ignored by MATLAB,
  but acted on and checked by LabMate. We outline the design
  principles behind LabMate, our current progress, and our future
  plans.
\end{abstract}

\begin{keywords}
  MATLAB, software correctness, dimensional consistency, type theory
\end{keywords}

@@

\[
\begin{pmatrix*}
  d_{\textrm{nc}1} & d_{\textrm{c}1} & n_1 & c_1 & t_1  \\
  d_{\textrm{nc}2} & d_{\textrm{c}2} & n_2 & c_2 & t_2  \\
   & \vdots &  & & \\
  d_{\textrm{nc}N} & d_{\textrm{c}N} & n_N & c_N & t_N  \\
\end{pmatrix*}
\times
\begin{pmatrix}
  a \\
  a_{cal} \\
  c \\
  c_{cal} \\
  m
\end{pmatrix}
=
\begin{pmatrix}
  v_1 \\
  v_2 \\
  \vdots \\
  v_N
\end{pmatrix}
\]
%
The matrix left of $\times$ is made up of the following values:
%
\begin{itemize}
    \item The values in the first four columns are purely numeric and have no associated dimension.

@@

    \item The fourth column is the opposite of the third.

@@

The column vector right of $\times$ and left of $=$ contains values to
be calculated using the line fit which then determine the measured
value of the resistor by a calculation which is beyond the scope of
this paper. The column vector right of $=$ contains voltage readings.

This calculation can be converted into a MATLAB problem using the
builtin left division operator \texttt{\textbackslash} to compute a
least squares fit. Let us step through how we intend to use LabMate to
help us deliver this solution.

We then ask for the type of the resulting solution. Note
that for reasons of space only twelve values are listed. A real
measurement would contain considerably more values:
\begin{verbatim}
%> z3 :: [ 3 x 1 ] double
z3 = [ 0; 0; 0 ]
%> i3 :: [ 3 x 1 ] double
i3 = [ 1; 1; 1 ]
%> times :: [ 12 x 1 ] double
% define times as a column vector
  of time measurements

ddnc = [ i3  z3  i3  z3
        -i3  z3  i3  z3
         z3  i3  z3  i3
         z3 -i3  z3  i3 ]
A = [ ddnc times ]
%> typeof A
%< A :: [ 12 x 5 ] double

@@

Did you notice the error in the above code? The vector \texttt{b} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate detects this error and produces the following as output.

%%> A :: [ 12 x 5 ] double
%A = [ 1  0  1  0  t1
%      1  0  1  0  t2
%      1  0  1  0  t3
%     -1  0  1  0  t4
%     -1  0  1  0  t5
%     -1  0  1  0  t6
%      0  1  0  1  t7
%      0  1  0  1  t8
%      0  1  0  1  t9
%      0 -1  0  1  t10
%      0 -1  0  1  t11
%      0 -1  0  1  t12 ]

\begin{verbatim}

%> z3 :: [ 3 x 1 ] double
z3 = [0; 0; 0]
%> i3 :: [ 3 x 1 ] double
i3 = [1; 1; 1]
%> times :: [ 12 x 1 ] double
ddnc = [ i3  z3  i3  z3
        -i3  z3  i3  z3
         z3  i3  z3  i3
         z3 -i3  z3  i3 ]
A = [ ddnc times ]
%> typeof A
%< A :: [ 12 x 5 ] double

@@
